
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reproducible Research : Course Project 1

## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Data for the Analysis

Dataset: activity.csv

The variables included in this dataset are:

. steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

. date: The date on which the measurement was taken in YYYY-MM-DD format

. interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

Load the data 

```{r}
library(csv)
activity<-read.csv("activity.csv")
```

Process/transform the data (if necessary) into a format suitable for your analysis

```{r}
activity$DateTime<- as.POSIXct(activity$date, format="%Y-%m-%d")
activity$day <- weekdays(as.Date(activity$date))
#pulling data without NAS
activityclean <- activity[!is.na(activity$steps),]
```

## What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

. Calculate the total number of steps taken per day

```{r}
sumTable <- aggregate(activity$steps ~ activity$date, FUN=sum)
colnames(sumTable)<- c("Date", "Steps")
sumTable
```

```{r}
## Creating the historgram of total steps per day
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day")
```

. Calculate and report the mean and median of the total number of steps taken per day

```{r}
## Mean of Steps
as.integer(mean(sumTable$Steps))
```

```{r}
## Median of Steps
as.integer(median(sumTable$Steps))
```

## What is the average daily activity pattern?

. Make a time series plot (i.e. type = "1") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
library(plyr)
library(ggplot2)
##pulling data without nas
activityclean <- activity[!is.na(activity$steps),]

##create average number of steps per interval
stepsperintervalTable <- ddply(activityclean, .(interval), summarize, Avg = mean(steps))

##Create line plot of average number of steps per interval
p <- ggplot(stepsperintervalTable, aes(x=interval, y=Avg), xlab = "Interval", ylab="Average Number of Steps")
p + geom_line()+xlab("Interval")+ylab("Average Number of Steps")+ggtitle("Average Number of Steps per Interval")
```

. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
##Maximum steps by interval
maxSteps <- max(stepsperintervalTable$Avg)
##Which interval contains the maximum average number of steps
stepsperintervalTable[stepsperintervalTable$Avg==maxSteps,1]
```

## Imputing missing values

Note that there are a number of days/intervals where there are missing values NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
##Number of NAs in original data set
nrow(activity[is.na(activity$steps),])
```

. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Strategy: to fill in all the NAs, I will substitute the missing steps with the average 5-minute interval, based on the days of the week.

```{r}
## Create the average number of steps per weekday and interval
avgTable <- ddply(activityclean, .(interval, day), summarize, Avg = mean(steps))

## Create dataset with all NAs for substitution
nadata<- activity[is.na(activity$steps),]
## Merge NA data with average weekday interval for substitution
newdata<-merge(nadata, avgTable, by=c("interval", "day"))
```

. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
## Reorder the new substituded data in the same format as clean data set
newdata2<- newdata[,c(6,4,1,2,5)]
colnames(newdata2)<- c("steps", "date", "interval", "day", "DateTime")

##Merge the NA averages and non NA data together
mergeData <- rbind(activityclean, newdata2)
```

. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 


```{r}
##Create sum of steps per date to compare with step 1
sumTable2 <- aggregate(mergeData$steps ~ mergeData$date, FUN=sum)
colnames(sumTable2)<- c("Date", "Steps")
```

```{r}
## Mean of Steps with NA data taken care of
as.integer(mean(sumTable2$Steps))
```

```{r}
## Median of Steps with NA data taken care of
as.integer(median(sumTable2$Steps))
```

```{r}
## Creating the histogram of total steps per day, categorized by data set to show impact
hist(sumTable2$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="Blue")
hist(sumTable$Steps, breaks=5, xlab="Steps", main = "Total Steps per Day with NAs Fixed", col="grey", add=T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("blue", "grey") )
```

Do these values differ from the estimates from the first part of the assignment? 

Yes. The new mean of the imputed data is 10821 steps compared to the old mean of 10766 steps. That creates a difference of 55 steps on average per day, producing a 0.5% error which is very low.

What is the impact of imputing missing data on the estimates of the total daily number of steps?

The new median of the imputed data is 11015 steps compared to the old median of 10765 steps. That creates a difference of 250 steps for the median producing an error of 2.6% which is very low, that is why the overall shape of the distribution has not changed.

## Are there differences in activity patterns between weekdays and weekends?

. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
## Create new category based on the days of the week
mergeData$DayCategory <- ifelse(mergeData$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
```

. Make a panel plot containing a time series plot (i.e.type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
library(lattice) 
## Summarize data by interval and type of day
intervalTable2 <- ddply(mergeData, .(interval, DayCategory), summarize, Avg = mean(steps))
##Plot data in a panel plot
xyplot(Avg~interval|DayCategory, data=intervalTable2, type="l",  layout = c(1,2),
       main="Average Steps per Interval Based on Type of Day", 
       ylab="Average Number of Steps", xlab="Interval")
```       

Answer: Yes. There are differences in activity patterns between weekdays and weekends. 
 
 
       